\sys: Light-touch Data-Parallel Shell Processing

### 2020-10-06 paper/draft.tex
Pass over section 5 and comments

After translating the dataflow regions of the input script to DFGs and applying the transformations~\sx{ir:transformations},
  \sys translates all DFGs back into a shell script.
Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
% The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.
% \kk{I removed this, since we describe them just below.}

@@

\section{Runtime Library}
% FIXME two labels here
\label{impl}
\label{optimizer}

@@

This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's custom primitives.
%% it instantiates the \todo{custom} split and relay nodes as highly
%% optimized commands to improve In this section we describe how \sys
%% achieves that, and several technical challenges that it addresses.

@@

\heading{Overcoming Laziness}
The shell's evaluation strategy is unusually lazy, in that most commands and shell constructs consume their inputs only when they are ready to process more.
% This allows \unix to multiplex the executing commands efficiently by interleaving their executions. 
%% support concurrently executing many more commands than the
%% processing threads of the underlying system,
%% by scheduling them concurrently, allowing them to
%% have progress one at a time.
Such laziness inhibits opportunities for parallelism, as commands are often blocked when their consumers are not requesting any input.
Consider the following fragment:

@@

\heading{Dangling FIFOs and Zombie Producers}
%
%% \TODO{(This problem is a nice followup for the one above) Problem:
%%   Dataflow graph termination. Illustration: Show the example of head
%%   deadlock and explain that this is inherent behaviour in the shell.
%%   (Solution: emptying pipelines or terminating processes) In order for
%%   things to terminate gracefully, dish wraps all commands in order to
%%   always consume all of their input by sending it down to
%%   /dev/null. Warning: This means that all commands in the pipeline
%%   must terminate gracefully if all of their output is consumed (if a
%%   process dies unexpectedly, noone will consume its input. We leave
%%   addressing that and other monitoring questions for future work). We
%%   evaluate the overhead of emptying all outputs in section X.}
%
% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a command exits after it has produced and sent all its results to its output channel.
If the channel is a pipe and its reader exits early, the command is notified to stop writing early.
In \unix, this is achieved by an out-of-band error mechanism: the
operating system delivers a \ttt{PIPE} signal to the producer,
notifying it that the pipe's consumer has exited.  This is different
from errors for other system calls
%% return an error
and unusual compared to non-\unix systems\footnote{For example,
  Windows indicates errors for \ttt{WriteFile} using its return
  code---similar to \ttt{DeleteFile} and other Win32 functions.}
primarily because pipes and pipelines are at the heart of \unix.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet
% ---possible for named pipes---
  \unix cannot signal this condition.
% \kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}
% \nv{TODO! Will think about this later}
Consider the following script:
% emitted code from \sys's earlier versions:
%% \kk{Proposed alternative: ``Consider the following shell script:''}
% \begin{lstlisting}[language=sh, numbers=none]
%   mkfifo fifo1 fifo2
%   cat file-chunk1.txt > fifo1 &
%   cat file-chunk2.txt > fifo2 &
%   cat fifo1 fifo2 | head -n 1 & wait
% \end{lstlisting}
\begin{minted}[fontsize=\footnotesize]{bash}
  mkfifo fifo1 fifo2
  cat file-chunk1.txt > fifo1 &
  cat file-chunk2.txt > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{minted}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
This, in turn, leaves it unable to make progress, as it is both blocked and unaware of its consumer exiting.
Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.

>>>>

After translating the dataflow regions of the input script to DFGs and applying the transformations~\sx{ir:transformations},
  \sys translates all DFGs back into a shell script.
Nodes of the graph are instantiated with the commands and flags they represent, and edges are instantiated as named pipes.
A prologue in the script creates the necessary intermediate pipes, and a \ttt{trap} call takes care of cleaning things up in cases where the script aborts.
% \kk{What does this "trap" refer to?}
% The script makes use of \sys-provided primitives available in the system \ttt{PATH} and described next.
% \kk{I removed this, since we describe them just below.}

@@

\section{Runtime}
% FIXME two labels here
\label{impl}
\label{optimizer}

@@

This section describes technical challenges related to execution of the resulting script and how they are addressed by \sys's runtime and custom primitives.
%% it instantiates the \todo{custom} split and relay nodes as highly
%% optimized commands to improve In this section we describe how \sys
%% achieves that, and several technical challenges that it addresses.

@@

\heading{Overcoming Laziness}
% \kk{eurosys: Maybe we can call this section pull vs push, and rephrase the terminology to explain the problem as "execution is bounded by the speed in which commands pull their inputs, leading to underutilization". Using eager we get rid of this bound prefetch the inputs of the next command.}
The shell's evaluation strategy is unusually lazy, in that most commands and shell constructs consume their inputs only when they are ready to process more.
% This allows \unix to multiplex the executing commands efficiently by interleaving their executions. 
%% support concurrently executing many more commands than the
%% processing threads of the underlying system,
%% by scheduling them concurrently, allowing them to
%% have progress one at a time.
Such laziness leads to CPU underutilization, as commands are often blocked when their consumers are not requesting any input.
Consider the following fragment:

@@

\heading{Dangling FIFOs and Zombie Producers}
%
%% \TODO{(This problem is a nice followup for the one above) Problem:
%%   Dataflow graph termination. Illustration: Show the example of head
%%   deadlock and explain that this is inherent behaviour in the shell.
%%   (Solution: emptying pipelines or terminating processes) In order for
%%   things to terminate gracefully, dish wraps all commands in order to
%%   always consume all of their input by sending it down to
%%   /dev/null. Warning: This means that all commands in the pipeline
%%   must terminate gracefully if all of their output is consumed (if a
%%   process dies unexpectedly, noone will consume its input. We leave
%%   addressing that and other monitoring questions for future work). We
%%   evaluate the overhead of emptying all outputs in section X.}
%
% \nv{I think all of these titles should be the problem/challenge, not the solution}
% A process writing to a pipe needs to be notified when its consumer exits early before consuming all its output.
Under normal operation, a command exits after it has produced and sent all its results to its output channel.
If the channel is a pipe and its reader exits early, the command is notified to stop writing early.
In \unix, this is achieved by an out-of-band error mechanism: the
operating system delivers a \ttt{PIPE} signal to the producer,
notifying it that the pipe's consumer has exited.  This is different
from errors for other system calls
%% return an error
and unusual compared to non-\unix systems\footnote{For example,
  Windows indicates errors for \ttt{WriteFile} using its return
  code---similar to \ttt{DeleteFile} and other Win32 functions.}
primarily because pipes and pipelines are at the heart of \unix.
% write errors are communicated differently 
Unfortunately though, if a pipe has not been opened for writing yet
% ---possible for named pipes---
  \unix cannot signal this condition.
% \kk{I think that it is also possible for anonymous, e.g. \ttt{cat f1 - | head -n 1}}
% \nv{TODO! Will think about this later}
Consider the following script:
% emitted code from \sys's earlier versions:
%% \kk{Proposed alternative: ``Consider the following shell script:''}
% \begin{lstlisting}[language=sh, numbers=none]
%   mkfifo fifo1 fifo2
%   cat file-chunk1.txt > fifo1 &
%   cat file-chunk2.txt > fifo2 &
%   cat fifo1 fifo2 | head -n 1 & wait
% \end{lstlisting}
\begin{minted}[fontsize=\footnotesize]{bash}
  mkfifo fifo1 fifo2
  cat in1 > fifo1 & cat in2 > fifo2 &
  cat fifo1 fifo2 | head -n 1 & wait
\end{minted}
\noindent
In the code above, \ttt{head} exits early causing the last \ttt{cat} to exit before opening \ttt{fifo2}.
As a result, the second \ttt{cat} never receives a \ttt{PIPE} signal that its consumer exited---after all, \ttt{fifo2} never even had a consumer.
This, in turn, leaves it unable to make progress, as it is both blocked and unaware of its consumer exiting.
Coupled with \ttt{wait} at the end, the entire snippet reaches a deadlock.
